Open-source landscape for 3D CSEM modelling

### 2020-04-13 manuscript/Manuscript.tex
Draft Block Model

% Remove once done
\newcommand{\mycom}[2][]{
  \todo[]{\textbf{\uppercase{[#1]}}:~#2}}
\newcommand{\imycom}[2][]{
  \todo[inline]{\textbf{\uppercase{[#1]}}:~#2}}
\newcommand{\dul}[1]{\underline{\underline{#1}}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\rmk}[1]{{\color{red}{#1}}}

@@

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft;\\
\footnotemark[2]TODO\\
E-mail: \href{mailto:Dieter@Werthmuller.org}{Dieter@Werthmuller.org};\\
\textbf{Keywords}: CSEM, Open-Source, 3D modelling, FD, FE.
}

\author{%
Dieter Werthmüller\footnotemark[1], %            orcid: 0000-0002-8575-2484
Raphael Rochlitz\footnotemark[2], %              orcid: 0000-0002-5132-916X
Octavio Castillo-Reyes\footnotemark[2], and %    orcid: 0000-0003-4271-5015
Lindsey Heagy\footnotemark[2] %                  orcid: 0000-0002-1551-5926
}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
\emph{A will draft a first abstract after the manuscript has been reviewed by
everyone once.}
%
% % Principal objectives and scope of the work
% % Methodology
% % Results
% % Conclusions
%
\end{abstract}

@@

In the following a quick rundown of the codes. It is, however, beyond the scope
of this article to go into every detail of the different modellers, and we
refer to their documentations for more details. An overview comparison of the
different codes is given in Tabel~\ref{tbl:codecomparison}. All codes have in
common that they solve the Maxwell's equation in its differential form (DE)
under the quasistatic or diffusive approximation, hence neglecting displacement
currents.

\tabl[btp]{codecomparison}{All codes solve Maxwell's equations in its
  differential form (DE) under the quasistatic approximation. Note that
  \emg3d is a solver on its own, while the other codes implement third-package
  solvers such as PETSc, MUMPS, FEniCS \citep{CSE.15.Alnaes}, or PARDISO
  \citep{FGCS.04.Schenk}.
  \imycom[DW]{PETSc and MUMPS probably need citations too.}}{
  \centering
  \footnotesize
\begin{tabular}{lcccc}
  \toprule
  %
  & \custem & \emg3d & \petgem & \simpeg  \\
  \midrule
  %
  Home & \href{https://custem.rtfd.io}{custem.rtfd.io}
       & \href{https://empymod.github.io}{empymod.github.io}
       & \href{http://petgem.bsc.es}{petgem.bsc.es}
       & \href{https://docs.simpeg.xyz}{simpeg.xyz} \\
  License & GPL-3.0 & Apache-2.0 & GPL-3.0 & MIT \\
  Installation & \texttt{conda}
               & \texttt{pip}; \texttt{conda}
               & \texttt{pip}; \texttt{conda}
               & \texttt{pip}; \texttt{conda} \\
  Comp. Dom. & frequency & frequency & frequency & frequency \& time \\
  Method & FE & FIT & FE & FV \\
  Mesh & tetrahedral & rectilinear & tetrahedral & rectilinear; curvilinear \\
  BC & Dirichlet & PEC & Neumann; Dirichlet & Dirichlet; Neumann \\
  Solver & PETSc; MUMPS & \emg3d & FEniCS & PARDISO; MUMPS \\
  %
  \bottomrule
\end{tabular}}%
%

@@

@@

In this numerical section compute the responses for two different models to
verify that these different 3D codes yield the same electromagnetic responses.
The first model is a simple one consisting of a layered, anisotropic
background, in which three blocks are embedded, and the second model is a very
complex, realistic model.

@@

\cite{GJI.13.Miensopust} published a review of two workshops dealing with the
verification of magnetotelluric forward and inversion codes, hence a similar
exercise to what we do here for CSEM codes. The block model is a derivation of
the \emph{Dublin Test Model 1}, a model from their first workshop. We took the
same layout of the blocks but adjusted the dimensions and resistivities to a
typical, marine CSEM problem, as shown in Figure~\ref{fig:block-model}. We
additionally added a layered background with vertical transverse isotropy
(VTI).

\imycom[DW]{Next:\\
  - Describe models in text\\
  - Write text for layered and block model\\
  - Start tables for them\\
  - Move to Marlim\\
  - Discussion\\
  - Conclusions\\
  - Send to Raphael
}

%
\plot*{block-model}{width=.6\textwidth}{
  Sketch of the block model. The layered model consists of an air layer, a
  water layer, a thin top-layer and then a thick background layer, and at the
  bottom a resistive basement layer. All three blocks are embedded in the thick
  background layer, which has VTI with $\lambda=\sqrt{2}$.
}
%

% %
% \tabl[btp]{comp1}{Caption.}{
%   \centering
% \begin{tabular}{lccrS[table-format=2.4]rl}
%   \toprule
%   %
%   Code & Method & Mesh & \#Cells & {CPU} & RAM & Computer \\
%   \midrule
%   %
%   \custem & FE  & FE & & & & \\
%   \emg3d  & FIT & FD & & & & \\
%   \petgem & FE  & FE & & & & \\
%   \simpeg & FD  & OcTree & & & & \\
%   %
%   \bottomrule
% \end{tabular}}%
% %

%
\plot*{results-layered-0}{width=\textwidth}{
   Comparison of the layered model at y=0km.
}
%

%
\plot*{results-layered--3}{width=\textwidth}{
   Comparison of the layered model at y=-3km.
}
%

%
\plot*{results-block--3}{width=\textwidth}{
   Comparison of the block model at y=-3km.
}
%

%
\plot*{results-block-0}{width=\textwidth}{
   Comparison of the block model at y=0km.
}
%

%
\plot*{results-block-3}{width=\textwidth}{
   Comparison of the block model at y=+3km.
}
%

@@

@@

>>>>

% Remove once done
\newcommand{\mycom}[2][]{
  \todo[]{\textbf{\uppercase{[#1]}}:~#2}}
\newcommand{\imycom}[2][]{
  \todo[inline]{\textbf{\uppercase{[#1]}}:~#2}}
\newcommand{\dul}[1]{\underline{\underline{#1}}}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\rmk}[1]{{\color{red}{#1}}}
\newcommand{\ohmm}{\ensuremath{\Omega\,}\text{m}\xspace}

@@

\address{
\footnotemark[1]TU Delft,
Building 23,
Stevinweg 1 / PO-box 5048,
2628 CN Delft;\\
\footnotemark[2]TODO\\
\footnotemark[3]TODO\\
\footnotemark[4]TODO\\
E-mail: \href{mailto:Dieter@Werthmuller.org}{Dieter@Werthmuller.org};\\
\textbf{Keywords}: CSEM, Open-Source, 3D modelling, FD, FE.
}

\author{%
Dieter Werthmüller\footnotemark[1], %            orcid: 0000-0002-8575-2484
Raphael Rochlitz\footnotemark[2], %              orcid: 0000-0002-5132-916X
Octavio Castillo-Reyes\footnotemark[3], and %    orcid: 0000-0003-4271-5015
Lindsey Heagy\footnotemark[4] %                  orcid: 0000-0002-1551-5926
}

@@

%%fakesection ===    ABSTRACT    ===
\begin{abstract}
%
\emph{I will draft a first abstract after the manuscript has been reviewed by
everyone once.}
%
% % Principal objectives and scope of the work
% % Methodology
% % Results
% % Conclusions
%
\end{abstract}

@@

In the following a quick rundown of the codes. It is, however, beyond the scope
of this article to go into every detail of the different modellers, and we
refer to their documentations for more details. An overview comparison of the
different codes is given in Table~\ref{tbl:codecomparison}. All codes have in
common that they solve the Maxwell's equation in its differential form (DE)
under the quasistatic or diffusive approximation, hence neglecting displacement
currents. The machines on which the different codes were run are listed in
Table~\ref{tbl:machines}, together with the responsible operator.

\tabl[btp]{codecomparison}{All codes solve Maxwell's equations in its
  differential form (DE) under the quasistatic approximation. Note that
  \emg3d is a solver on its own, while the other codes implement third-package
  solvers such as \texttt{PETSc}, \texttt{MUMPS}, \texttt{FEniCS}
  \citep{CSE.15.Alnaes}, or \texttt{PARDISO} \citep{FGCS.04.Schenk}.
  \imycom[DW]{\texttt{PETSc} and \texttt{MUMPS} probably need citations too.}}{
  \centering
  \footnotesize
\begin{tabularx}{\linewidth}{lXXXX}
  \toprule
  %
  & \custem & \emg3d & \petgem & \simpeg  \\
  \midrule
  %
  Home & \href{https://custem.rtfd.io}{custem.rtfd.io}
       & \href{https://empymod.github.io}{empymod.github.io}
       & \href{http://petgem.bsc.es}{petgem.bsc.es}
       & \href{https://docs.simpeg.xyz}{simpeg.xyz} \\
  License & GPL-3.0 & Apache-2.0 & GPL-3.0 & MIT \\
  Installation & \texttt{conda}
               & \texttt{pip}; \texttt{conda}
               & \texttt{pip}; \texttt{conda}
               & \texttt{pip}; \texttt{conda} \\
  Comp. Dom. & frequency & frequency & frequency & frequency \& time \\
  Method & FE & FIT & FE & FV \\
  Mesh & tetrahedral & rectilinear & tetrahedral & recti- \& curvilinear \\
  BC & PEC & PEC & Neumann; Dirichlet & Dirichlet; Neumann \\
  Solver & \texttt{PETSc}; \texttt{MUMPS} & \emg3d & \texttt{FEniCS} &
           \texttt{PARDISO}; \texttt{MUMPS} \\
  %
  \bottomrule
\end{tabularx}}%
%

@@

%
\tabl[btp]{machines}{List of computer and operating system on which the
  different codes were run, together with the operator.}{
  \centering
  \footnotesize
  \begin{tabularx}{\linewidth}{lXl}
  \toprule
  %
  Code & Computer and Operating System & Operator \\
  \midrule
  %
  \custem & PowerEdge R940 server; 144 Xeon Gold 6154 CPU @2.666 GHz; ~3 TB
            DDR4 RAM; Ubuntu 18.04
          & Raphael Rochlitz \\
  \emg3d  & Laptop; i7-6600U CPU@2.6 GHz x4; 16 GB of memory, Ubuntu 18.04
          & Dieter Werthmüller \\
  \petgem & Marenostrum4. Intel Xeon Platinum from Skylake generation; 2
            sockets Intel Xeon Platinum 8160 CPU with 24 cores each @2.10GHz
            for a total of 48 cores per node; 386 Gb DDR4 RAM per node; SuSE
            Linux Enterprise
          & Octavio Castillo-Reyes \\
  \simpeg & ???
          & Lindsey Heagy \\
  %
  \bottomrule
\end{tabularx}}%
%

@@

In this numerical section we compute the responses for two different models to
verify that these different 3D codes yield the same electromagnetic responses.
The first model is a simple one consisting of a layered, anisotropic
background, in which three blocks are embedded, and the second model is a very
complex, realistic model.

@@

\cite{GJI.13.Miensopust} published a review of two workshops dealing with the
verification of magnetotelluric forward and inversion codes, hence a similar
exercise to what we do here for CSEM codes. The block model is a derivation of
the \emph{Dublin Test Model 1}, a model from their first workshop. We took the
same layout of the blocks but adjusted the dimensions and resistivities to a
typical, marine CSEM problem, as shown in Figure~\ref{fig:block-model}. We
additionally added a layered background with vertical transverse isotropy
(VTI).
%
\plot*{block-model}{width=.6\textwidth}{
  Sketch of the block model. The layered model consists of an air layer, a
  water layer, a thin top-layer and then a thick background layer, and at the
  bottom a resistive basement layer. All three blocks are embedded in the thick
  background layer, which has VTI with $\lambda=\sqrt{2}$.
}
%

The layered model consists of an upper halfspace of air, a 600\,m deep water
layer, followed by a 150\,m thick, isotropic layer of 1\,\ohmm, a 3.3\,km
thick, anisotropic layer of $\rho_\text{h}=2\,\ohmm$ and
$\rho_\text{v}=4\,\ohmm$, and finally a resistive basement consisting of a
lower halfspace of $1000\,\ohmm$. The 200\,m long, $x$-directed source is
located 50\,m above the seafloor from $x=-50$ to $x=50\,$m in $x$-direction, at
$y=0\,$m. The $x$-directed receivers are placed on the seafloor every 200\,m
from $x=-10\,$km to $x=+10\,$km in three lines for $y=-3, 0, 3\,$km.

In a first step we compare the layered background to the semi-analytical
solutions of a 1D code, for which we use \empymod \citep{GEO.17.Werthmuller}.
Figures~\ref{fig:results-layered-0} and \ref{fig:results-layered--3} show the
actual responses in the top rows, and the relative error in the bottom rows
for receiver line $y=0\,$km and $y=-3\,$km, respectively (for the layered
background the responses for the two receiver-lines of $y=\pm3$\,km are
identical, so we only show one here).

Figure~\ref{fig:results-layered-0} shows in the top-row the inline responses
for offsets $r>0.5\,$km. As all codes compute the same we only show the
semi-analytical result for the real and imaginary amplitudes, and then in the
bottom row the relative error of the 3D codes. The results show that the codes
are literally yielding the same result, the relative error is for big parts
below 1\,\%, and for almost all parts below a few percent. Some differences are
due to the chosen discretizations. E.g., \emg3d is less precise close to the
200\,m long source. This could be improved by choosing smaller cell sizes, at
the cost of computation time. At large offsets, on the other hand, \emg3d seems
to do a very good job.
%
\plot*{results-layered-0}{width=.8\textwidth}{
   Comparison of the layered model for the inline receivers ($y=0\,$km).
   Top row shows the semi-analytical responses from \empymod, and bottom row
   show the relative error (\%) of \custem, \emg3d, \petgem, and \simpeg.
}
%

The corresponding broadside responses for $y=-3\,$km is shown in
Figure~\ref{fig:results-layered--3}, and the required CPU and RAM is shown in
Table~\ref{tbl:comp-layered}.
%
\plot*{results-layered--3}{width=.8\textwidth}{
   Comparison of the layered model for the broadside receivers ($y=-3\,$km).
   Top row shows the semi-analytical responses from \empymod, and bottom row
   show the relative error (\%) of \custem, \emg3d, \petgem, and \simpeg.
}
%

%
\tabl[btp]{comp-layered}{Comparison of used CPU and RAM and of number of cells
  used to discretize the layered model. \imycom[DW]{I will fill this table out
  once we all submitted the final versions of the results.}}{
  \centering
\begin{tabular}{lS[table-format=4.2]rS[table-format=4.2]rl}
  \toprule
  %
  Code & {CPU} & \#Procs & RAM     & \#Cells \\
       & {(s)} &         & {(GiB)} &         \\
  \midrule
  %
  \custem & 0.0 & 24 & 0.0 & 0 \\
  \emg3d  & 0.0 &  1 & 0.0 & 0 \\
  \petgem & 0.0 & 48 & 0.0 & 0 \\
  \simpeg & 0.0 &    & 0.0 & 0 \\
  %
  \bottomrule
\end{tabular}}%
%

Three resistive blocks are added in the 3.3\,km-thick, anisotropic background
layer, as shown in Figure~\ref{fig:block-model}. They have resistivities of
$\rho=10\,\ohmm$ (shallow beam perpendicular to survey lines),
$\rho=100\,\ohmm$ (thin plate, South-East), $\rho=500\,\ohmm$ (cube,
North-West). As there are no analytical solutions for this model we show the
normalized difference between different codes, instead of the relative error,
where the normalized difference of two responses $p$ and $q$ is given by
%
\begin{equation}
  \mr{NRMSD~(\%)} = 200 \frac{|p - q|}{|p| + |q|}\ .
  \label{eq:nrmsd}
\end{equation}
%
\imycom[DW]{NRMSD: Comparing all codes to all would make the
figures too cluttered, and I don't think it would add a lot insights. My idea
is therefore to compare \emg3d/\simpeg (the two FD ones), \custem/\petgem (the
two FE ones), and \emg3d/\custem (a cross-comparison). What do you think?}

The results for the three receiver lines $y=-3,0,3\,$km are shown in
Figures~\ref{fig:results-block--3}, \ref{fig:results-block-0}, and
\ref{fig:results-block-3}, respectively, and the corresponding CPU and RAM in
Table~\ref{tbl:comp-block}. What we can conclude from the NRMSD is that similar
codes have a smaller NRMSD than cross-comparison, hence \emg3d and \simpeg are
very similar, as well as are \custem and \petgem.
\mycom[DW]{Write in detail when all results are available}
%
\plot*{results-block--3}{width=.8\textwidth}{
   Comparison of the block model for the broadside receivers at $y=-3\,$km.
   Top row shows the responses and bottom row the NRMSDs (\%)
   between \emg3d--\custem, \custem--\petgem, and \emg3d--\simpeg.
}
%

%
\plot*{results-block-0}{width=.8\textwidth}{
   Comparison of the block model for the inline receivers at $y=0\,$km.
   Top row shows the responses and bottom row the NRMSDs (\%)
   between \emg3d--\custem, \custem--\petgem, and \emg3d--\simpeg.
}
%

%
\plot*{results-block-3}{width=.8\textwidth}{
   Comparison of the block model for the broadside receivers at $y=+3\,$km.
   Top row shows the responses and bottom row the NRMSDs (\%) between
   \emg3d--\custem, \custem--\petgem, and \emg3d--\simpeg.
}
%

%
\tabl[btp]{comp-block}{Comparison of used CPU and RAM and of number of cells
  used to discretize the block model. \imycom[DW]{I will fill this table out
  once we all submitted the final versions of the results.}}{
  \centering
\begin{tabular}{lS[table-format=4.2]rS[table-format=4.2]rl}
  \toprule
  %
  Code & {CPU} & \#Procs & RAM     & \#Cells \\
       & {(s)} &         & {(GiB)} &         \\
  \midrule
  %
  \custem & 0.0 & 24 & 0.0 & 0 \\
  \emg3d  & 0.0 &  1 & 0.0 & 0 \\
  \petgem & 0.0 & 48 & 0.0 & 0 \\
  \simpeg & 0.0 &    & 0.0 & 0 \\
  %
  \bottomrule
\end{tabular}}%
%

@@

\imycom[DW]{Next:\\
  - Create figures for Marlim\\
  - Write text for Marlim\\
  - Discussion\\
  - Conclusions\\
  - Send to Raphael
}

@@

%
\tabl[btp]{comp-marlim}{Comparison of used CPU and RAM and of number of cells
  used to discretize the Marlim R3D model. \imycom[DW]{I will fill this table
  out once we all submitted the final versions of the results.}}{
  \centering
\begin{tabular}{lS[table-format=4.2]rS[table-format=4.2]rl}
  \toprule
  %
  Code & {CPU} & \#Procs & RAM     & \#Cells \\
       & {(s)} &         & {(GiB)} &         \\
  \midrule
  %
  \custem & 0.0 & 24 & 0.0 & 0 \\
  \emg3d  & 0.0 &  1 & 0.0 & 0 \\
  \petgem & 0.0 & 48 & 0.0 & 0 \\
  \simpeg & 0.0 &    & 0.0 & 0 \\
  %
  \bottomrule
\end{tabular}}%
%
